---
title: "Causal Map Functions Vignette"
author: "Steve Powell"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: false
    theme: lumen
editor_options: 
  chunk_output_type: console
---

# Causal Map 3 Functions

This file briefly illustrates the causal mapping functions for Causal Map 3.

The functions are used within Causal Map 3, which provides a web interface for managing causal mapping project and also enables autocoding with AI. The functions can also be used in the R console as illustrated here.  

These functions completely replace the functions in the old R package CausalMapFunctions which is now deprecated. This new set of functions is much simpler. 

#  Setup

```{r setup, cache=F,include=F}
# library(DBI)
# library(dbplyr)
# library(pool)
# library(dplyr, warn.conflicts = FALSE)
# library(httr)
 library(tidyverse)
# library(readxl)
# library(stringi)
# library(jsonlite)
 library(DiagrammeR)
# library(openxlsx)
knitr::opts_chunk$set(echo = T,warning=F,message=F)
```

# Load the functions and some example data

```{r setup2, cache=F,include=F}
library(causalmap3functions)

# source the functions and get example data
# for the convenience of the developer...
if(Sys.getenv("USERDOMAIN") == "STEVE-P") {
source("R/cm3functions.R") 
ex="example-file.xlsx" %>%
  get_from_excel()
  
} else {
  
source("https://gist.githubusercontent.com/stevepowell99/ae16b31779d1e462e5a3c7003eef79a1/raw/cm3functions.R")
url2 <- "https://figshare.com/ndownloader/files/42600379"
ex <- 
list(
  files=read.xlsx(url2,1),
  links=read.xlsx(url2,2),
  statements=read.xlsx(url2,3),
  sources=read.xlsx(url2,4)
)
} 

# constants -----------------------------------------------------------------

contrary_color <- "#f2ab73"
ordinary_color <- "#058488"
flipchar <- "~"



```

# Inspecting the example data file

```{r}
ex$links %>% 
  arrange(desc(source_count)) %>% 
  select(from_label,to_label,source_count,source_id,link_id,statement_id) %>% 
  head

ex$links %>% 
  arrange(desc(source_count)) %>% 
  distinct(from_label,to_label,.keep_all = T) %>% 
  select(from_label,to_label,source_count,source_id,link_id,statement_id) %>% 
  head

ex$links %>% 
  make_factors_from_links() %>% 
  arrange(desc(source_count)) %>% 
  head

ex$sources %>% 
  select(1:6) %>% 
  head

ex$statements %>% 
  select(1:6) %>% 
  head



```

# Filtering the links

These functions use tidyverse conventions and workflows throughout. 

There are no special functions for filtering the links because the tidyverse workflow is so simple:

```{r}
ex$links %>% 
  filter(str_detect(source_id,"MNX"))
ex$links %>% 
  filter(statement_id=="MNY-1 | 43")

```

However there is a special function for including and excluding factors by label.


# All the transforms functions

## Zoooming

```{r}
ex$links %>% 
  make_factors_from_links() %>% 
  arrange(desc(source_count)) %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  make_factors_from_links() %>% 
  arrange(desc(source_count)) %>% 
  head

```


## Selecting top links and factors

```{r}

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Sources",which="Top",top=3) %>%  
  make_print_map2(ex$links)

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Sources",which="Top",top=3) %>%  
  select(from_label,to_label,link_count) %>% 
  distinct() %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Sources",which="At_least",top=15) %>% 
  make_factors_from_links() %>% 
  arrange(desc(source_count)) %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Links",which="Top",top=12) %>%  
  make_factors_from_links() %>% 
  arrange(desc(link_count)) %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Links",which="At_least",top=20) %>%  
  make_factors_from_links() %>% 
  arrange(desc(link_count)) %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_links(type="Sources",which="Top",top=2) %>%  
  select(from_label,to_label,link_count) %>% 
  distinct() %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_links(type="Sources",which="At_least",top=4) %>%  
  select(from_label,to_label,link_count,source_count) %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_links(type="Links",which="Top",top=20) %>%  
  select(from_label,to_label,link_count,source_count) %>% 
  arrange(desc(link_count)) %>% 
  distinct() %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_links(type="Sources",which="Top",top=20) %>%  
  select(from_label,to_label,link_count,source_count) %>% 
  arrange(desc(source_count)) %>% 
  distinct() %>% 
  head

```

## Tracing paths

### Tracing paths, showing source IDs
```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 2,transforms_tracing_threads = F) %>%  
  pipe_label("source_id") %>% 
  make_print_map2(original = ex$links,map_color_factors_column = "outcomeness")
```



### Tracing threads, showing source IDs

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 2,transforms_tracing_threads = T) %>%  
  pipe_label("source_id") %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")

```

### Tracing paths, showing source counts

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 2,transforms_tracing_threads = F) %>%  
  pipe_label("source_count") %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")
```



### Tracing threads, showing source counts

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 2,transforms_tracing_threads = T) %>%  
  pipe_label("source_count") %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")

```

### Tracing just one step

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 1) %>%  
  make_print_map2(original = ex$links)
```

### Tracing threads, three steps

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_trace(sess_links = ex$links,from_labels = "Farm production",to_labels="Diet improved",steps = 3,transforms_tracing_threads = T) %>%  
  pipe_label("link_count") %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")
```


## Removing brackets

This file doesn't actually have factor names with brackets, but we can create some

```{r}
tmp <- 
  ex$links %>% 
  mutate(from_label=str_replace_all(from_label,"Wellbeing","Wellbeing (outcome)")) %>% 
  mutate(to_label=str_replace_all(to_label,"Wellbeing","Wellbeing (outcome)"))

tmp %>% 
  pipe_zoom() %>%
  pipe_focus(c("Wellbeing"),any=T) %>%  
  pipe_top_factors(5) %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")

tmp %>% 
  pipe_zoom() %>%
  pipe_focus(c("Wellbeing"),any=T) %>%  
  pipe_top_factors(5) %>% 
  pipe_remove_brackets(round = T) %>% 
  make_print_map2(original = ex$links,map_label_factors = "source_count")


```


## Focus

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_focus(c("Ability to buy food")) %>%  
  make_print_map2(original = ex$links,map_label_factors = "source_count")
  
ex$links %>% 
  pipe_zoom() %>%
  pipe_focus(c("Ability"),any = T) %>%  
  make_print_map2(original = ex$links,map_label_factors = "source_count")

ex$links %>% 
  pipe_zoom() %>%
  pipe_focus(c("Ability to buy food","Diet improved"),any = T) %>%  
  make_print_map2(original = ex$links,map_label_factors = "source_count")

ex$links %>% 
  pipe_zoom() %>%
  pipe_focus(c("Ability","Diet improved"),any = T) %>%  
  make_print_map2(original = ex$links,map_label_factors = "source_count")

```

## Exclude

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Links",which="Top",top=6) %>%
  select(from_label,to_label,link_count,source_count) %>% 
  distinct %>% 
  head

ex$links %>% 
  pipe_zoom() %>%
  pipe_top_factors(type="Links",which="Top",top=6) %>%
  pipe_exclude("Increased knowledge" %>% c("Diet improved")) %>% 
  select(from_label,to_label,link_count,source_count) %>% 
  distinct %>% 
  head

```



## Hashtags

```{r}
ex$links %>% 
  pipe_zoom() %>%
  pipe_retain_hashtags(hashtags = "#hypothetical",keep=T)  %>% 
  select(from_label,to_label,link_count,source_count,hashtags) 

ex$links %>% 
  pipe_zoom() %>%
  pipe_retain_hashtags(hashtags = "#hypothetical",keep=T)  %>% 
  pipe_label("hashtags") %>% 
  make_print_map2(ex$links)

ex$links %>% 
  pipe_zoom() %>%
  pipe_retain_hashtags(hashtags = "#hypothetical",keep=F)  %>% 
  filter(hashtags!="") %>% 
  select(from_label,to_label,link_count,source_count,hashtags) 


```


## Combining opposites

```{r}
ex$links %>% 
  pipe_focus(xc("~"),any = T) %>% 
  pipe_top_factors(20) %>% 
  make_print_map2(ex$links)

ex$links %>% 
  pipe_focus(xc("~"),any = T) %>% 
  pipe_top_factors(20) %>% 
  make_print_map2(ex$links,map_colour_opposites_red = T)

ex$links %>% 
  pipe_focus(xc("~"),any = T) %>% 
  pipe_combine_opposites() %>% 
  pipe_top_factors(20) %>% 
  make_print_map2(ex$links)


if(F){
  ex2 <- get_file("opposites-example",conn=conn)
  
ex2$links %>% 
  pipe_combine_opposites() %>% 
  make_print_map2(ex$links)
  
  
}


```

# Creating additional tables

## Mentions table

```{r}
ex$links %>% 
  make_mentions_tabl() %>% 
  select(label,direction,link_id)

ex$links %>% 
  make_mentions_tabl() %>% 
  select(label,direction,link_id) %>% 
  pivot_wider(names_from=direction,values_from = link_id,values_fn = length)

```


# Formatting the map

Note that labelling links is a pipe function, not part of the call to the map

```{r}
ex$links %>% 
  pipe_zoom(1) %>% 
  pipe_top_factors(8) %>% 
  make_print_map2(ex$links,map_color_factors_column = "outcomeness")

ex$links %>% 
  pipe_zoom(1) %>% 
  pipe_top_factors(8) %>% 
  pipe_label("source_count") %>% 
  make_print_map2(ex$links,map_color_factors_column = "source_count",map_size_factors = "source_count",map_label_factors = "source_count",map_size_links = "source_count",map_wrap_factor_labels = 9)

```


## Surprises

Calculating surprises is technically a formatting function

```{r}
ex$links %>% 
  pipe_zoom(1) %>% 
  pipe_top_links(8) %>% 
  pipe_label(map_label_links = "#Name of village" ,type="Surprise_links") %>% 
  make_print_map2(original = ex$links,map_color_factors_column = "link_count",map_size_links ="link_count")

ex$links %>% 
  pipe_zoom(1) %>% 
  pipe_top_links(8) %>% 
  pipe_label(map_label_links = "#Name of village",type="Surprise_sources" ) %>% 
  make_print_map2(original = ex$links,map_color_factors_column = "link_count",map_size_links ="link_count")




ex$links %>% 
  pipe_zoom(1) %>% 
  pipe_top_links(18) %>% 
  pipe_label(map_label_links = "#Name of village",type = "Surprise_links" ) %>% 
  make_print_map2(original = ex$links,map_color_factors_column = "link_count",map_size_links ="source_count")

```



